---
title: "Dimensional Reduction in R"
author: "Sudipta Hazra"
date: "`r Sys.Date()`"
mainfont: Times New Roman
output:
  md_document
#   html_document:
#     toc: true
#     toc_float: true
#     # number_section: true
# always_allow_html: yes
---

```{r setup}
proj = rprojroot::find_rstudio_root_file()
knitr:: opts_knit$set(root.dir = proj)

knitr::opts_chunk$set(
	fig.align = "center",
	message = FALSE,
	warning = FALSE,
	dpi = 600
)
```

# ***Dimensional Reduction Techniques for Proteomics & Chemometrics: PCA, t-SNE and UMAP***

*Sudipta Hazra*

*Abstract:*

This project is a part of **Food Proteomics Project**. It covers R code for different dimensional reduction techniques like,

a.  Principal Component Analysis (PCA)

b.  t-distributed Stochastic Neighbor Embedding (t-SNE)

c.  Uniform Manifold Approximation and Projection (UMAP)


## Introduction

### First time running the code,
The structure of the project folder is,

1. Download the whole project folder,

2. run a `renv::restore()`,

3. The folder structure is as below,
```{r echo=FALSE}
# Install fs package if needed
# install.packages("fs")

# Display up to 2 levels of the directory structure
fs::dir_tree(path = ".", recurse = 2)

```

4. Keep the data at `data/sample_data.csv`, the data must contain `sample`, `label` columns along with the numeric variables. Below is an example,

```{r echo=FALSE}
library(dplyr)
library(knitr)

# Number of proteins (e.g., n = 10 for illustration)
n <- 10

# Generate dynamic protein column names: "Protein 1", "Protein 2", ..., "Protein n"
protein_cols <- paste0("Protein ", 1:n)

# Generate random protein values for 3 samples
protein_values <- replicate(n, sample(10:30, 3, replace = TRUE))  # 3 rows of random protein values

# Create the dataframe with sample and label columns
data <- data.frame(
  sample = c("Sample1", "Sample2", "Sample3"),
  label = c("Label1", "Label2", "Label3")
)

# Add dynamically created protein columns to the dataframe
data <- cbind(data, setNames(as.data.frame(protein_values), protein_cols))

# Select the first and last few protein columns, and insert "..." in between
selected_data <- data %>%
  select(sample, label, `Protein 1`, `Protein 2`) %>%  # Select first few columns
  mutate("..." = "...") %>%  # Insert a column of ellipsis
  bind_cols(data %>% select(`Protein 9`, `Protein 10`))  # Bind the last few columns

# Create the markdown table using kable
kable(selected_data, format = "markdown")


```

5. The codes are at `scripts/02_Dimensional_Reduction.Rmd`, run it,

6. The plots will appear at `plots/`.


## Libraries

Libraries need to run the function are,

```{r}
# pacman is a wrapper function, it checkcs if a package is already installed then use library(), if doesn't exist then uses install.packages()
# install.packages("pacman")

# basic libraries
pacman::p_load(tidyverse, 
               ggplot2, 
               patchwork,
               plotly, # for interactive plotting
               glue,
               tsne, # for tsne
               umap, # for umap
               tidylog
               )

#for PCA
# basic stat package and ggplot is enough.
```

## Importing Data

```{r}
# importing data
sample_data  <- read.csv("data/sample_data.csv")

# separating numeric data
df <-  sample_data %>% 
  select(-"label")%>% 
  column_to_rownames("sample") 

# separating label (character/factor) data
labels <- sample_data %>% 
  select(c("sample", "label"))%>% 
  column_to_rownames("sample")
```

## General Functions

1.  General plot function to plot the PCA, t-SNE and UMAP result.

```{r}

#general Dimension reduction Plot (for PCA, t-SNE, UMAP a general scatter plot function)

generate_dimension_reduction_plot <- function(model, x, y, model_output_data, label, labels, explained_variance = NULL) {
  
  # pca data and corresponding "sample" name
  model_output_data <- model_output_data  %>%
    rownames_to_column("sample")
  
  # group label and corresponding "sample" name
  # label <- "Feeding_Regime"
  group_names <- labels %>% 
    select(all_of(label))%>% 
    rownames_to_column("sample")
  
  colnames(group_names)[2] <- "group" 
  
  # joining pca data and group label by "sample"
  model_output_data <- model_output_data %>% 
    full_join(group_names, by = "sample") %>% 
    column_to_rownames("sample")
  
  
  print(model_output_data)
  
  p <-
    ggplot(model_output_data,
           aes(x = model_output_data[, x], y = model_output_data[, y], color = group)) +
    geom_point(size = 4, alpha = 5 / 10) +  # Marker size
    geom_hline(yintercept = 0, linetype="dashed", alpha = 3/10) +
    geom_vline(xintercept = 0, linetype="dashed", alpha = 3/10) +
    geom_polygon(stat = "ellipse",
                 aes(fill = group),
                 alpha = 0.01) +
    theme_bw(base_family = "serif")
  
  if (tolower(model) == "pca") {
    p <- p +
      labs(
        x = glue("Principal Component {x} ({explained_variance[x]}%)"),
        y = glue("Principal Component {y} ({explained_variance[y]}%)")
      )
    
  } else if (tolower(model) == "tsne"){
    p <- p  +
      labs(
        x = glue("tSNE {x}"),
        y = glue("tSNE {y}")
      )
  } else if (tolower(model) == "umap"){
    p <- p  +
      labs(
        x = glue("UMAP {x}"),
        y = glue("UMAP {y}")
      )
  } else{
    p <- p  +
      labs(
        x = glue("Component {x}"),
        y = glue("Component {y}")
      )
  }
  
  return(p)
}
```

2.  Function for plotting the PCA scree plot.

```{r}
generate_pca_scree_plot <- function(scree_plot_data, n){
  scree_plot <- ggplot(scree_plot_data[1:n,], mapping = aes(x = reorder(PC_component, explained_variance), y = explained_variance)) +
    geom_bar(stat = "identity", fill="#56B4E9", colour="black") +
    geom_line(aes(x = 1:n, y = explained_variance), linewidth = 1) +
    geom_text(aes(label = round(explained_variance, 2)), vjust = -0.5, size = 3)+
    labs(x = "Principal Componenets",
         y = "Explained Variance (%)") +
    theme_bw(base_family = "serif")
  return(scree_plot)
}

```

## Dimensionaity Reduction

### Principal Component Analysis (PCA)

#### Defining the PCs to plot

```{r}
x = 1 # PC1
y = 2 # PC2
model = "pca"
```

#### Model building

```{r}
pca_model <- prcomp(df, scale. = TRUE)
```

#### Extracting the PCA results

```{r}
pca_data <- as.data.frame(pca_model$x)
```

#### Calculating explained variance

```{r}
explained_variance <- pca_model$sdev^2 / sum(pca_model$sdev^2)
explained_variance <- round(explained_variance * 100, 2)
```

#### Preparing Scree plot

```{r}
scree_plot_data <- data.frame(PC_component = paste0('PC ', 1: nrow(sample_data)), explained_variance = cumsum(explained_variance))
components_to_show = 10

scree_plot <- generate_pca_scree_plot(scree_plot_data, components_to_show)
```

#### Preparig Score plot

```{r}
for (label in colnames(labels)) {
  assign(paste0("fig_pca_scorePlot_", label), generate_dimension_reduction_plot("pca",x, y, pca_data, label, labels, explained_variance = explained_variance))
}
```

#### Plotting Score Plot and Scree Plot side by side

```{r}
fig_pca_scorePlot_label / scree_plot
```

### t-distributed Stochastic Neighbor Embedding (t-SNE)

#### Defining the parameters to plot

```{r}
x = 1
y = 2 
perplexity = 20
model = "tsne"
```

#### Model building

```{r}
set.seed(0)
tsne_model <- tsne(df, initial_dims = 2)
```

#### Extracting the t-SNE results

```{r}
tsne_data <- data.frame(tsne_model)
rownames(tsne_data) <- rownames(df)
```

#### Preparing t-SNE plots

```{r}
for (label in colnames(labels)) {
  assign(paste0("fig_tsne_Plot_", label), generate_dimension_reduction_plot(model, x, y, tsne_data, label, labels, explained_variance = NULL))
}
```

#### Plotting t-SNE plots

```{r}
fig_tsne_Plot_label
```

### Uniform Manifold Approximation and Projection (UMAP)

#### Defining parameters to plot

```{r}
x = 1
y = 2 
model = "umap"
```

#### Model building

```{r}
set.seed(0)
umap_model <- umap(df, n_components = 2, random_state = 15)
```

#### Extracting UMAP results

```{r}
umap_data <- data.frame(umap_model[["layout"]]) 
```

#### Preparing UMAP plots

```{r}
for (label in colnames(labels)) {
  assign(paste0("fig_umap_Plot_", label), generate_dimension_reduction_plot(model, x, y, umap_data, label, labels, explained_variance = NULL))
}

fig_umap_Plot_label
```

### Save all the plots

```{r}
items <- ls()

for (item in items){
  # print(item)
  if(grepl("fig", item)){
    ggsave(get(item), file = glue("plots/{item}.png"), dpi = 600)
  }
}
```
